---
title: "A hands-on tutorial on R and RStudio: worked solutions"
author: "Tiago A. Marques"
date: today
format:
  html:
    embed-resources: true
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
    fig-width: 9
    fig-height: 5.5
  pdf:
    toc: true
    number-sections: true
    fig-width: 9
    fig-height: 5.5
  docx:
    toc: true
    number-sections: true
    fig-width: 9
    fig-height: 5.5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
old_options <- options(width = 90, digits = 4)
```

# How to use this report

This is a complete worked solution to the activities in
`TAMsIntro2RviaRStudioTutorial.qmd`. The emphasis is not only on obtaining an
answer, but also on making each analysis reproducible. Run the chunks in order;
several later chunks use objects created earlier.

There is rarely only one correct way to solve an R problem. The solutions below
use mostly base R so that the ideas remain visible. A reproducible report should
contain the data import, analysis, figures, numerical results, and interpretation—not
just the final answer.

# Dynamic reports and reproducibility

The separate reproducibility example asks us to repeat an analysis after the data
change from four to six habitats. The essential edit is simply:

```{r}
#| eval: false
mydata <- read.csv("Examples/DynamicReproductibleDataAnalysis/fakedata2.csv")
```

All summaries, plots, and fitted models downstream then update when the document
is rendered. This is the central lesson: code and prose remain tied to the data,
so a revised data file does not require hand-copying results into a report.

# Task 0: objects, calculations, and cleanup

```{r}
# Create three numeric objects
length_m <- 12
width_m  <- 5
height_m <- 2

# Use them in calculations
area_m2   <- length_m * width_m
volume_m3 <- length_m * width_m * height_m
c(area_m2 = area_m2, volume_m3 = volume_m3)

# Remove the temporary objects created in this section
rm(length_m, width_m, height_m, area_m2, volume_m3)
```

The assignment operator `<-` stores values under names. Arithmetic with numeric
objects works just as it does with numbers. `rm()` removes named objects; in a
real analysis, it is usually clearer to restart R and rerun the report than to
depend on manual workspace cleanup.

## Interpreting coercion

```{r}
obj1 <- factor(c("a", "a", "b", "a", "e", "f", "d"))
as.integer(c(3, -0.3, 0.4, 0.6, 0.9, 13.2, 12))
as.numeric(c(TRUE, FALSE, TRUE))
as.numeric(obj1)
```

`as.integer()` truncates the fractional part toward zero. Logical values become
1 (`TRUE`) and 0 (`FALSE`). Converting a factor directly to numeric returns its
internal level codes, **not** the printed labels; that is why factor coercion
must be handled carefully.

# Task 1: mathematical and statistical functions

We use small vectors so every result can be checked by hand.

```{r}
x <- c(1, 2, 3, 4, 5)
y <- c(2, 1, 4, 3, 7)

c(sum = sum(x), square_root_9 = sqrt(9), natural_log_e = log(exp(1)))
c(log_base_10_1000 = log(1000, base = 10), exp_2 = exp(2))
c(combinations_5_choose_2 = choose(5, 2), factorial_5 = factorial(5))
c(floor = floor(2.8), ceiling = ceiling(2.2), rounded = round(pi, 3))
c(absolute = abs(-7), cos_0 = cos(0), sin_pi_over_2 = sin(pi / 2))
c(tan_pi_over_4 = tan(pi / 4), acos_1 = acos(1), acosh_1 = acosh(1))
c(max = max(x), min = min(x), mean = mean(x), median = median(x))
range(x)
var(x)
cor(x, y)
quantile(x)
```

The first group transforms values or performs combinatorial calculations. The
second summarizes a vector. `var()` is the sample variance (denominator
`length(x) - 1`), `cor()` measures linear association, and `quantile()` returns
selected cut-points of a distribution.

# Task 2: extending R with packages

As a useful ecological example, the `vegan` package provides tools for community
ecology, including diversity indices and ordination. Installation is performed
once at the console, not while rendering a report:

```{r}
#| eval: false
install.packages("vegan")       # run once, in the Console
library(vegan)                  # run in each new R session/report
data(dune)
diversity(dune, index = "shannon")
```

The optional `memer` challenge is shown without executing it because it requires
a GitHub installation and external image assets:

```{r}
#| eval: false
install.packages("remotes")
remotes::install_github("sctyner/memer")
library(memer)
meme_get("DistractedBf") |>
  meme_text_distbf("tidyverse", "new R users", "base R")
```

The distinction is important: install a package once; load it with `library()`
whenever a session needs it.

# Task 3: importing `dados1.csv`

```{r}
newfile <- read.csv("dados1.csv")
head(newfile)
str(newfile)
summary(newfile)
```

The file has `r nrow(newfile)` rows and `r ncol(newfile)` columns. `read.csv()`
is appropriate because the first row contains column names and fields are
comma-separated. Inspecting `str()` immediately after import is good practice:
it confirms that the columns have the expected classes.

# Task 4: plotting the `cars` data

```{r}
data(cars)
special <- cars$speed < 15 & cars$dist > 70

plot(dist ~ speed, data = cars,
     xlab = "Speed (mph)", ylab = "Stopping distance (ft)",
     pch = 16, col = "grey35",
     main = "Stopping distance as a function of speed")
points(cars$speed[special], cars$dist[special],
       pch = 21, cex = 1.7, lwd = 2, col = "firebrick", bg = "gold")
text(cars$speed[special], cars$dist[special],
     labels = which(special), pos = 3, col = "firebrick")
legend("topleft", legend = "speed < 15 mph and distance > 70 ft",
       pch = 21, pt.bg = "gold", col = "firebrick", bty = "n")
```

In this dataset, `r sum(special)` observation(s) meet both conditions. Notice
that the original `cars` documentation reports distance in **feet**, not metres.

# Task 5: regression and prediction at 16 mph

```{r}
mylm1 <- lm(dist ~ speed, data = cars)
pred16 <- unname(predict(mylm1, newdata = data.frame(speed = 16)))

plot(dist ~ speed, data = cars,
     xlab = "Speed (mph)", ylab = "Stopping distance (ft)",
     xlim = c(0, 30), ylim = c(0, 120), pch = 16, col = "grey40")
abline(mylm1, lwd = 2)
abline(v = 16, lty = 2, col = "royalblue")
abline(h = pred16, lty = 2, col = "royalblue")
points(16, pred16, pch = 19, col = "royalblue", cex = 1.3)
text(16, pred16, labels = paste0("  predicted = ", round(pred16, 1), " ft"),
     pos = 4, col = "royalblue")

pred16
```

The fitted linear model predicts a stopping distance of approximately
`r round(pred16, 1)` ft at 16 mph. This is a mean-response prediction, not a
guarantee for an individual car. Extrapolation to very low speeds exposes a
limitation of the Gaussian straight-line model: it can predict impossible
negative distances.

# Task 6: simulation and probability distributions

## Normal samples of two sizes

```{r}
set.seed(606)
mu <- 2
sigma <- 1.5
x50 <- rnorm(50, mean = mu, sd = sigma)
x5000 <- rnorm(5000, mean = mu, sd = sigma)

par(mfrow = c(1, 2))
hist(x50, probability = TRUE, breaks = "FD", col = "grey85", border = "white",
     xlim = c(-4, 8), main = "n = 50", xlab = "Simulated value")
curve(dnorm(x, mean = mu, sd = sigma), add = TRUE,
      col = "firebrick", lwd = 2)
hist(x5000, probability = TRUE, breaks = "FD", col = "grey85", border = "white",
     xlim = c(-4, 8), main = "n = 5000", xlab = "Simulated value")
curve(dnorm(x, mean = mu, sd = sigma), add = TRUE,
      col = "firebrick", lwd = 2)
par(mfrow = c(1, 1))
```

Both samples come from the same distribution, but the histogram based on 5,000
observations follows the theoretical density much more closely. Larger samples
reduce Monte Carlo noise; they do not change the population distribution.

## Three beta distributions

```{r}
curve(dbeta(x, 1, 1), from = 0, to = 1, ylim = c(0, 5), lwd = 2,
      xlab = "x", ylab = "Density", main = "Selected beta distributions")
curve(dbeta(x, 1, 5), add = TRUE, col = "royalblue", lwd = 2)
curve(dbeta(x, 0.5, 0.5), add = TRUE, col = "firebrick", lwd = 2,
      n = 1001)
legend("topright", c("Beta(1,1)", "Beta(1,5)", "Beta(0.5,0.5)"),
       col = c("black", "royalblue", "firebrick"), lwd = 2, bty = "n")
```

Beta(1,1) is uniform. Beta(1,5) concentrates near zero. Beta(0.5,0.5) is
U-shaped and is often called the **bathtub-shaped** beta distribution. Because
beta variables lie between 0 and 1, they are useful for proportions and
probabilities, such as occupancy, survival, or detection probabilities.

# Task 7: conditional logic inside a loop

```{r}
set.seed(707)
par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))
sample_means <- numeric(9)

for (i in seq_len(9)) {
  z <- rnorm(40)
  sample_means[i] <- mean(z)
  line_colour <- if (sample_means[i] >= 0) "royalblue" else "firebrick"

  hist(z, breaks = 10, col = "grey85", border = "white",
       main = paste("Sample", i), xlab = "Value")
  abline(v = sample_means[i], lty = 2, lwd = 2, col = line_colour)
}
par(mfrow = c(1, 1))

round(sample_means, 3)
```

The loop repeats the same operation nine times. The `if` statement chooses blue
for a non-negative sample mean and red for a negative one. Even though the true
mean is zero, finite samples fluctuate around zero.

# Task 8: writing functions

```{r}
mystats <- function(x) {
  if (!is.numeric(x)) stop("x must be numeric")
  if (length(x) == 0L) stop("x must contain at least one value")

  x_complete <- x[!is.na(x)]
  non_negative <- x_complete[x_complete >= 0]

  c(
    mean = mean(x_complete),
    variance = var(x_complete),
    maximum = max(x_complete),
    minimum = min(x_complete),
    mean_without_negative = if (length(non_negative)) mean(non_negative) else NA_real_
  )
}

mystats(c(-4, -1, 0, 3, 7, NA))
```

Here “excluding negative numbers” means retaining zero and positive values. The
function also deals explicitly with missing values and gives informative errors.

A second useful function standardizes a numeric vector (subtract the mean and
divide by the standard deviation):

```{r}
standardize <- function(x) {
  if (!is.numeric(x)) stop("x must be numeric")
  s <- sd(x, na.rm = TRUE)
  if (is.na(s) || s == 0) stop("x must have a positive standard deviation")
  (x - mean(x, na.rm = TRUE)) / s
}

round(standardize(c(10, 12, 15, 20)), 3)
```

# Capstone: a distance-sampling survey

## Design and an important geometric consequence

The stated study area is 10 km long and 1 km wide. A longitudinal transect can
be placed at a random position across the width. With a 500 m truncation distance,
a transect at the exact centre covers the entire 1 km width. A transect elsewhere
would leave an edge strip outside the covered region. To honor the exercise while
avoiding that asymmetry, each simulated transect is placed uniformly and the
animal x-coordinates are measured relative to it with periodic (wrap-around)
boundaries. This is equivalent to joining the two width edges and removes edge
effects without changing the stated dimensions.

```{r}
N <- 10000
study_length <- 10000 # metres
study_width  <- 1000  # metres
w <- 500              # truncation distance (metres)
sigma_det <- 200       # half-normal scale (metres)

half_normal <- function(distance, sigma = 200) {
  exp(-(distance^2) / (2 * sigma^2))
}

wrapped_distance <- function(x, transect_x, width) {
  direct <- abs(x - transect_x)
  pmin(direct, width - direct)
}
```

## One survey

```{r}
set.seed(808)
animals <- data.frame(
  x = runif(N, 0, study_width),
  y = runif(N, 0, study_length)
)
transect_x <- runif(1, 0, study_width)
animals$distance <- wrapped_distance(animals$x, transect_x, study_width)
animals$available <- animals$distance <= w
animals$p_detect <- half_normal(animals$distance, sigma_det)
animals$detected <- animals$available & runif(N) < animals$p_detect

c(available = sum(animals$available), detected = sum(animals$detected))
```

All `r sum(animals$available)` animals are within 500 m under this design (the
largest possible wrapped perpendicular distance is 500 m), but detection remains
imperfect and declines with distance.

```{r}
plot(animals$x, animals$y, pch = ".", col = "grey70",
     xlab = "Position across study area (m)", ylab = "Position along study area (m)",
     main = "Simulated population and detections")
abline(v = transect_x, col = "royalblue", lwd = 2)
points(animals$x[animals$detected], animals$y[animals$detected],
       pch = 16, cex = 0.45, col = adjustcolor("firebrick", 0.7))
legend("topright", c("transect", "detected animal"),
       col = c("royalblue", "firebrick"), lwd = c(2, NA),
       pch = c(NA, 16), bty = "n")
```

## Detection function and empirical detection probability

```{r}
distance_grid <- seq(0, w, length.out = 501)
plot(distance_grid, half_normal(distance_grid, sigma_det), type = "l", lwd = 3,
     xlab = "Perpendicular distance (m)", ylab = "Detection probability",
     ylim = c(0, 1), main = "Half-normal detection function")
```

```{r}
breaks <- seq(0, w, by = 50)
distance_bin <- cut(animals$distance, breaks = breaks, include.lowest = TRUE)
empirical_p <- tapply(animals$detected, distance_bin, mean)
bin_midpoints <- head(breaks, -1) + diff(breaks) / 2

plot(bin_midpoints, empirical_p, pch = 19, ylim = c(0, 1),
     xlab = "Perpendicular distance (m)", ylab = "Proportion detected",
     main = "Observed and theoretical detection probability")
lines(distance_grid, half_normal(distance_grid, sigma_det),
      lwd = 2, col = "royalblue")
legend("topright", c("observed by 50 m bin", "half-normal model"),
       pch = c(19, NA), lty = c(NA, 1), col = c("black", "royalblue"), bty = "n")
```

The average detection probability, conditional on being within 500 m, is the
area under the detection curve divided by 500:

```{r}
true_p <- integrate(half_normal, lower = 0, upper = w,
                    sigma = sigma_det)$value / w
observed_p <- mean(animals$detected[animals$available])
c(theoretical = true_p, observed_this_survey = observed_p)
```

Thus the theoretical probability is about `r round(true_p, 3)`. With 10,000
animals we expect roughly `r round(N * true_p)` detections per complete survey.

## Repeat the complete survey 500 times

```{r}
simulate_survey <- function(N = 10000, length = 10000, width = 1000,
                            w = 500, sigma = 200) {
  x <- runif(N, 0, width)
  y <- runif(N, 0, length) # simulated because it is part of the spatial population
  transect_x <- runif(1, 0, width)
  distance <- wrapped_distance(x, transect_x, width)
  available <- distance <= w
  detected <- available & runif(N) < half_normal(distance, sigma)

  c(available = sum(available), detected = sum(detected))
}

set.seed(809)
survey_results <- replicate(
  500,
  simulate_survey(N, study_length, study_width, w, sigma_det)
)
number_detected <- survey_results["detected", ]
summary(number_detected)
sd(number_detected)
```

```{r}
hist(number_detected, breaks = "FD", col = "skyblue3", border = "white",
     xlab = "Number of detected animals", main = "Results from 500 complete surveys")
abline(v = mean(number_detected), col = "firebrick", lwd = 2, lty = 2)
abline(v = N * true_p, col = "black", lwd = 2, lty = 3)
legend("topright", c("simulated mean", "theoretical expectation"),
       col = c("firebrick", "black"), lty = c(2, 3), lwd = 2, bty = "n")
```

The repeated surveys vary because both animal positions and detection events are
new each time. The distribution is centred close to the theoretical expectation,
showing that the simulation and analytical calculation agree. The exercise also
demonstrates why raw counts underestimate abundance when detection is imperfect.

# Capstone: a permutation test for sequential occurrences

The tutorial text contains an inconsistency: it first asks whether squirrels
appear after **weasels**, but later states a badger hypothesis and counts the
reverse sequence `squirel.badger`. Here we answer the first, clearest scientific
question:

> Do squirrels occur immediately after weasels more often than expected if the
> observed species labels were randomly ordered?

The null hypothesis is that sequence order is unrelated to species identity,
conditional on the observed abundance of every species. The statistic is the
number of adjacent `weasel -> squirel` pairs.

```{r}
set.seed(189)
species <- c("wolf", "bear", "deer", "racoon", "weasel", "bobcat",
             "badger", "squirel", "sheep")
abundances <- c(50, 20, 300, 500, 500, 50, 400, 1000, 200)
appearences <- rep(species, times = abundances)
appearences <- sample(appearences, replace = FALSE)

count_transition <- function(sequence, from, to) {
  sum(head(sequence, -1) == from & tail(sequence, -1) == to)
}

observed_stat <- count_transition(appearences, "weasel", "squirel")
observed_stat
```

We repeatedly permute the complete sequence. This preserves species counts but
destroys any serial association.

```{r}
set.seed(190)
B <- 9999
null_stats <- replicate(
  B,
  count_transition(sample(appearences, replace = FALSE), "weasel", "squirel")
)

# One-sided test: unusually many weasel -> squirrel transitions
p_value <- (1 + sum(null_stats >= observed_stat)) / (B + 1)
c(observed = observed_stat,
  null_mean = mean(null_stats),
  p_value = p_value)
```

```{r}
hist(null_stats, breaks = seq(min(null_stats) - 0.5, max(null_stats) + 0.5, by = 1),
     col = "grey80", border = "white",
     xlab = "Weasel followed by squirrel transitions",
     main = "Permutation distribution under the null hypothesis")
abline(v = observed_stat, col = "firebrick", lwd = 3)
legend("topright", paste("observed =", observed_stat),
       col = "firebrick", lwd = 3, bty = "n")
```

The one-sided Monte Carlo p-value is `r format.pval(p_value, digits = 3)`. A small
p-value would mean the observed count is unusually high under random ordering;
a large p-value would mean the observed count is compatible with the null model.
This test supports an association in sequence, not necessarily a causal effect.

# Final checklist

A strong student report should:

1. keep all analysis code in executable chunks;
2. use relative paths and inspect data immediately after import;
3. label figures with variables and units;
4. use `set.seed()` before stochastic work when exact reproducibility matters;
5. distinguish sample results from population or model quantities;
6. state assumptions and interpret results in words; and
7. render from a fresh R session to ensure no hidden workspace objects are needed.

```{r}
#| label: restore-options
#| include: false
options(old_options)
```
